Dark matter web in the Universe

### 2018-06-15 Draft/Chapter4/Paper2017a_Nobmath.tex
bruce

The rest of the paper is organized as follows: we describe the cosmological simulations in Section \ref{sec:simulation}. Some of the important features of the multistream field are described in Section \ref{sec:multiCalc}. Topology of the single-streaming voids is discussed in \ref{sec:voids} and that of the multistream structure is investigated using percolation theory in Section \ref{sec:percolation}. . Discussion of the local geometry of multistream field using Hessian matrices is done in Section \ref{sec:hessian}.

@@

% \bf changed to \bf --- check if it's alright. 
The multistream field objectively characterizes the level of non-linearity in the cosmic web. The `number-of-streams' field or $n_{str}({\bf x})$ is computed from the Lagrangian sub-manifold ${\bf x}({\bf q})$, which is a continuous three-dimensional sheet in a six-dimensional  $({\bf q}, {\bf x})$ space. In this paper, we utilize the tessellation implementation by \cite{Shandarin2012} to calculate the multistream flow field on the GADGET-2 snapshot at $z=0$. This implementation only requires initial and final coordinates of the dark matter particles.

@@

The excursion volume fraction of the multistream web structure is limited to a small fraction of  less than 10 per cent since rest of the volume is void. The excursion set volume fraction increases with decreasing number-of-streams and reaches it's maximum at $n_{str} = 3$.  At this limit the filling fraction $f_1/f_{ES}$ is still less than unity, about 95 per cent. These two peculiar properties of the multistream field explain the shape of the percolation curves in \autoref{fig:PercTh}.
Since the multistream flow field is a discrete data field, the percolation transition is seen to occurs at a particular value of $n_{str}$ rather than a large range of values. For $n_{str} = 17$, the largest structure in the excursion set occupies more than half the volume of the entire excursion set. At this multistream threshold, the largest segment starts spanning large volume of the simulation box (as observed in the left panel of \autoref{fig:HaloFil}). The volume fraction of the excursion set at this percolation transition is $f_{ES}^{(p)} = 0.75$ per cent for simulation with $256^3$ particles.

@@

Smoother multistream fields result in less noisy PDFs of the Hessian eigenvalues. For instance, the volume fraction of regions with positive curvature (i.e. $\lambda_3 > 0$) is $2.4\%$, $2.3\%$ and $2.5\%$ for scales $0.20 h^{-1} \text{ Mpc}$, $0.39 h^{-1} \text{ Mpc}$, $0.78 h^{-1} \text{ Mpc}$ respectively. Further analysis of smoothed positive definite regions is relevant in determining halo boundaries, and will be extensively discussed in the next paper.

>>>>

The rest of the chapter is organized as follows: we describe the cosmological simulations in Section \ref{sec:simulation}. Some of the important features of the multistream field are described in Section \ref{sec:multiCalc}. Topology of the single-streaming voids is discussed in \ref{sec:voids} and that of the multistream structure is investigated using percolation theory in Section \ref{sec:percolation}. . Discussion of the local geometry of multistream field using Hessian matrices is done in Section \ref{sec:hessian}.

@@

% \bf changed to \bf --- check if it's alright. 
The multistream field objectively characterizes the level of non-linearity in the cosmic web. The `number-of-streams' field or $n_{str}({\bf x})$ is computed from the Lagrangian sub-manifold ${\bf x}({\bf q})$, which is a continuous three-dimensional sheet in a six-dimensional  $({\bf q}, {\bf x})$ space. In this chapter, we utilize the tessellation implementation by \cite{Shandarin2012} to calculate the multistream flow field on the GADGET-2 snapshot at $z=0$. This implementation only requires initial and final coordinates of the dark matter particles.

@@

The excursion volume fraction of the multistream web structure is limited to a small fraction of  less than 10 per cent since rest of the volume is void. The excursion set volume fraction increases with decreasing number-of-streams and reaches it's maximum at $n_{str} = 3$.  At this limit the filling fraction $f_1/f_{ES}$ is still less than unity, about 95 per cent. These two peculiar properties of the multistream field explain the shape of the percolation curves in \autoref{fig:PercTh}.
Since the multistream flow field is a discrete data field, the percolation transition is seen to occur at a particular value of $n_{str}$ rather than a large range of values. For $n_{str} = 17$, the largest structure in the excursion set occupies more than half the volume of the entire excursion set. At this multistream threshold, the largest segment starts spanning large volume of the simulation box (as observed in the left panel of \autoref{fig:HaloFil}). The volume fraction of the excursion set at this percolation transition is $f_{ES}^{(p)} = 0.75$ per cent for simulation with $256^3$ particles.

@@

Smoother multistream fields result in less noisy PDFs of the Hessian eigenvalues. For instance, the volume fraction of regions with positive curvature (i.e. $\lambda_3 > 0$) is $2.4\%$, $2.3\%$ and $2.5\%$ for scales $0.20 h^{-1} \text{ Mpc}$, $0.39 h^{-1} \text{ Mpc}$, $0.78 h^{-1} \text{ Mpc}$ respectively. Further analysis of smoothed positive definite regions is relevant in determining halo boundaries, and will be extensively discussed in the next chapter.

### 2018-06-15 Draft/Chapter6/concl.tex
bruce

Formation and evolution velocity multistreams in the dark matter Universe plays a vital role in the formation of large scale structures in the Universe. In this thesis, we have analyzed this multi-stream fields in cosmological context -- from dark matter halo environments to connectivities in the single streaming voids. It has been demonstrated that the studies of the Lagrangian sub-manifold not only provides a novel way of looking at the cosmos, but also boosts our current knowledge of the dark matter Universe.

@@

The full dynamical state of  dark matter can be described as a three-dimensional sub-manifold
in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
(where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis. Projecting the Lagrangian sub-manifold at each point in the configuration space defines the multistream field  $n_{str}(\mathbf{x}, z)$. This integer-valued field is equivalent to counting the number of foldings in the sub-manifold and we numerically calculate by a Lagrangian tessellation scheme. The resulting multistream field is dynamical, positively correlated with matter density, and supplements our knowledge of spatial clustering. Our major results can be summarized as follows.

@@

Voids are uniquely defined by the local condition requiring to be a single-stream flow region. Within the dark matter web (regions with $n_{str} > 1$), The walls are regions where number of streams is equal to three or a few. The number of streams in the neighboring filaments is higher than in the neighboring walls. Finally, each halo or sub-halo is a local peak in the $n_{str}$ field. We demonstrated in Chapter \ref{chapter3} that the shells of streams around haloes are quite thin and the closest void region is typically within one and a half FOF radius from the center of the halo.

@@

Rapid topological transitions from are seen in matter density fields as well as multistream fields, i.e., at a given threshold (matter density or multistream), the filling fraction of the largest structure in the Universe dominates the excursion volume fraction. For the multistreaming excursion sets, the percolating structure is significantly thinner than the filaments in over-density excursion approach. We also demonstrated that the matter density field is more fragmented than the multistream field.

@@

\begin{figure} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/f1fES_caustic.pdf} 
\caption{Percolation in caustic particles with variation of $l_{max}$: Top panel shows the mass fraction of the largest isolated caustic surface $f_1$ and mass fraction of all particles in the caustic $f_{ES}$. Bottom panel shows the filling fraction $f_1/f_{ES}$. Between 3.5-4 $h^{-1} Mpc$ the caustic surfaces transition into distinct isolated surfaces -- i.e., forming turn-around boundaries of haloes. }
\label{fig:caustic_perc_lmax}
\end{figure}

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{$L=20$ Mpc, $N_p = 32^3$. Small simulation, Caustic are filtered by longest side length of the triangle: Top left: $ l_{max} \geq 0$, Top right: $ l_{max} \geq 1$, Bottom left: $ l_{max} \geq 2$, Bottom right: $ l_{max} \geq 3$. }
\label{fig:caustic_lmax}
\end{figure*}

Despite the theoretical understanding mentioned above, delineating caustic surfaces at various levels of gravitational collapse has been proven to be difficult. Instead, a combination of geometric and topological methods may also enable us to trace the caustic formation. The DM particles on the caustics are on curved two dimensional surfaces, forming vertices of triangles that constitute the {\it tiles} of the surface. These triangles are of various shapes and size. Since the mean separation of particles is smaller than elsewhere in the cosmic web, the triangles are smaller on an average. Using the largest side of the triangle $l_{max}$ as a proxy for length scale of the triangle, one may perform morphological analyses on the type of caustic surfaces. One such study is shown in \autoref{fig:caustic_perc_lmax}. Percolation transition (in an unstructured grid, as opposed to regular grid analyses for multistream or flip-flop fields) are used to differentiate the larger caustic surface (like the ones between voids and walls) and the smaller ones (like the concentric caustic shells around haloes). This transition in our small cosmological simulation of side length $L = 20 h^{-1} Mpc$ is seen at around $3.5-4 h^{-1} Mpc$.

@@

Careful examination of the smaller caustic surfaces around the haloes may reveal concentric shells where the particles {\it turn back}. A spherical approximation of this, called the {\it splashback radius} \citep{More2015}. This has gained a considerable attention recently due to possible observational signatures of physical boundaries of dark matter haloes \citep{Chang2017}. Caustic finding scheme with appropriate filtering provides a general method to find these boundaries exactly. Moreover, mapping caustic surfaces from Eulerian space to Lagrangian space displays concentric surfaces, providing a dynamical insight into the splashback radii of dark matter haloes.

>>>>

Formation and evolution of velocity multistreams in the dark matter Universe plays a vital role in the formation of large scale structures in the Universe. In this thesis, we have analyzed this multi-stream fields in cosmological context -- from dark matter halo environments to connectivities in the single streaming voids. It has been demonstrated that the studies of the Lagrangian sub-manifold not only provides a novel way of looking at the cosmos, but also boosts our current knowledge of the dark matter Universe.

@@

The full dynamical state of cold dark matter can be described as a three-dimensional sub-manifold
in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
(where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis. Projecting the Lagrangian sub-manifold at each point in the configuration space defines the multistream field  $n_{str}(\mathbf{x}, z)$. This integer-valued field is equivalent to counting the number of foldings in the sub-manifold and we numerically calculate by a Lagrangian tessellation scheme. The resulting multistream field is dynamical, positively correlated with matter density, and supplements our knowledge of spatial clustering. Our major results can be summarized as follows.

@@

Voids are uniquely defined by the local condition requiring to be a single-stream flow region. Within the dark matter web (regions with $n_{str} > 1$), the walls are regions where number of streams is equal to three or a few. The number of streams in the neighboring filaments is higher than in the neighboring walls. Finally, each halo or sub-halo is a local peak in the $n_{str}$ field. We demonstrated in Chapter \ref{chapter3} that the shells of streams around haloes are quite thin and the closest void region is typically within one and a half FOF radius from the center of the halo.

@@

Rapid topological transitions are seen in matter density fields as well as multistream fields, i.e., at a given threshold (matter density or multistream), the filling fraction of the largest structure in the Universe dominates the excursion volume fraction. For the multistreaming excursion sets, the percolating structure is significantly thinner than the filaments in over-density excursion approach. We also demonstrated that the matter density field is more fragmented than the multistream field.

@@

\begin{figure} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/f1fES_caustic_new.pdf} 
\caption{Percolation in caustic particles with variation of $l_{max}$: Top panel shows the mass fraction of the largest isolated caustic surface $f_1$ and mass fraction of all particles in the caustic $f_{ES}$. Bottom panel shows the filling fraction $f_1/f_{ES}$. Between $1.1-1.4 h^{-1} Mpc$ the caustic surfaces transition into distinct isolated surfaces -- i.e., forming turn-around boundaries of haloes. }
\label{fig:caustic_perc_lmax}
\end{figure}

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{Caustic surfaces in a small cosmological simulation of $L=20$ Mpc, $N_p = 32^3$. Caustics are filtered by longest side length $l_{max}$ of the triangle consitituting the caustic surface. Top left: all values of $l_{max}$ (all the caustic surfaces), top right: $ l_{max} \leq 1.8  h^{-1} Mpc$, bottom left: $ l_{max} \leq 1.25  h^{-1} Mpc$, bottom right: $ l_{max} \leq 1.0  h^{-1} Mpc$.}
\label{fig:caustic_lmax}
\end{figure*}

Despite the theoretical understanding mentioned above, delineating caustic surfaces at various levels of gravitational collapse has been proven to be difficult. Instead, a combination of geometric and topological methods may also enable us to trace the caustic formation. The DM particles on the caustics are on curved two dimensional surfaces, forming vertices of triangles that constitute the {\it tiles} of the surface. These triangles are of various shapes and size. Since the mean separation of particles is smaller than elsewhere in the cosmic web, the triangles are smaller on an average. Using the largest side of the triangle $l_{max}$ as a proxy for length scale of the triangle, one may perform morphological analyses on the type of caustic surfaces. One such study is shown in \autoref{fig:caustic_perc_lmax}. Percolation transition (in an unstructured grid, as opposed to regular grid analyses for multistream or flip-flop fields) are used to differentiate the larger caustic surface (like the ones between voids and walls) and the smaller ones (like the concentric caustic shells around haloes). This transition in our small cosmological simulation of side length $L = 20 h^{-1} Mpc$ is seen at around $1.2 h^{-1} Mpc$.

@@

Careful examination of the smaller caustic surfaces around the haloes may reveal concentric shells where the particles {\it turn back}. A spherical approximation of this, called the {\it splashback radius} \citep{More2015}. This has gained a considerable attention recently due to possible observational signatures of physical boundaries of dark matter haloes \citep{Chang2017}. Caustic finding scheme with appropriate filtering provides a general method to find these boundaries in N-body simulations exactly. Moreover, mapping caustic surfaces from Eulerian space to Lagrangian space displays concentric surfaces, providing a dynamical insight into the splashback radii of dark matter haloes.

### 2018-06-15 Draft/thesis-ku.tex
thesis update
bruce

\dept{Department of Physics and Astronomy}
\degreetitle{Doctor of Philosophy}
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Hume Feldman}{Barbary Anthony-Twarog}{Mikhail Medvedev}{Suzanne Shontz}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web. Haloes are the highest concentrations of dark matter and host luminous galaxies. Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations.

% 2018 Paper
The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other.

@@

\include{Chapter1/intro}
\include{Chapter2/background}
\include{Chapter3/Paper2015}
\include{Chapter3b/Paper2018}
\include{Chapter4/Paper2017a_Nobmath}
\include{Chapter5/Paper2017b_Nobmath}
\include{Chapter6/concl}

>>>>

\dept{Department of Physics and Astronomy}
\degreetitle{Doctor of Philosophy}
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Barbary Anthony-Twarog}{Hume Feldman}{Suzanne Shontz}{Bruce Twarog}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

% Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web. Haloes are the highest concentrations of dark matter and host luminous galaxies. Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations.

% 2018 Paper
The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other. Using various geometrical, topological and heuristic tools in a novel parameter space -- the multistream field -- we explore the distribution of dark matter in cosmological simulations.

@@

\include{Chapter1/intro}
\include{Chapter2/background}
\include{Chapter3/Paper2015}
\include{Chapter4/Paper2017a_Nobmath}
\include{Chapter5/Paper2017b_Nobmath}
\include{Chapter3b/Paper2018}
\include{Chapter6/concl}
